Temporal semantics for a live coding language

### 2014-05-08 paper/paper.tex
some refactoring on exceptions

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

@@

\authorinfo{Sam Aaron}{}{}
\authorinfo{Dominic Orchard}{}{}
\title{Temporal semantics for a live coding language} 
% A programming model for temporal coordination (in music)}

@@

The \emph{Monad} instance for \emph{Temporal} is then as follows:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{T}\;\Varid{p})\bind \Varid{q}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
\\
\>[16]{}\hsindent{5}{}\<[21]%
\>[21]{}\mathbf{do}\;{}\<[25]%
\>[25]{}(\Varid{x},\Varid{vT'}){}\<[37]%
\>[37]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT}{}\<[E]%
\\
\>[25]{}\mathbf{let}\;(\Conid{T}\;\Varid{q'}){}\<[37]%
\>[37]{}\mathrel{=}\Varid{q}\;\Varid{x}{}\<[E]%
\\
\>[25]{}\Varid{thenT}{}\<[37]%
\>[37]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[25]{}\Varid{q'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
To ease understanding, we recall the types of \emph{return}
and \ensuremath{(\bind )} along with some intuition for their behaviour for
\emph{Temporal}:
%
\begin{itemize}
\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} lifts a pure value into a trivially
effectful computation by ignoring the time parameters and 
providing the usual pure state behaviour of returning the parameter state unchanged
(named \emph{vT} in this case).

@@

The following interpretation function $\interp{-}$ on \lang{} 
programs shows the mapping to the operations of the \emph{Temporal}
monad:
%%
\begin{align*}
\interp{\emph{statement}} & : \emph{Temporal} \, () \\
\interp{x = P; Q} & = \interp{P} \ensuremath{\bind (\lambda \Varid{x}\to } \interp{Q}) \\
\interp{P; Q} & = \interp{P} \ensuremath{\bind (\lambda \anonymous \to } \interp{Q}) \\
\interp{\sleep e} & = \emph{sleep} \, \interp{e}
\end{align*}
%%
Note that $\interp{-}$ is overloaded in the rule for \sleepOp{} for (pure) expressions. 
The concrete interpreation of other statements in the language, such as \playOp, is
elided here since it does not relate directly to the temporal semantics.

@@

\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{time}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
\\
\>[B]{}\Varid{time}{}\<[7]%
\>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
\\
\>[B]{}\Varid{start}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{startT},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{getVirtualTime}\mathbin{::}\Conid{Temporal}\;\Conid{VTime}{}\<[E]%
\\
\>[B]{}\Varid{getVirtualTime}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{vT},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{setVirtualTime}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
\\
\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\lambda \anonymous \to \lambda \anonymous \to \Varid{return}\;((),\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{kernelSleep}\mathbin{::}\Conid{RealFrac}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Temporal}\;(){}\<[E]%
\\
\>[B]{}\Varid{kernelSleep}\;\Varid{t}\mathrel{=}{}\<[18]%
\>[18]{}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to {}\<[E]%
\\
\>[18]{}\hsindent{6}{}\<[24]%
\>[24]{}\mathbf{do}\;{}\<[28]%
\>[28]{}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{t}\mathbin{*}\mathrm{1000000})){}\<[E]%
\\
\>[28]{}\Varid{return}\;((),\Varid{vT})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Simple \emph{Temporal} computations, used  by the model}
\label{core-functions}
\end{figure}

@@

\subsection{Emitting overrun exceptions}

>>>>

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

@@

\authorinfo{Sam Aaron}
           {Computer Laboratory, University of Cambridge, UK}
           {sam.saaron\ensuremath{\mathord{@}}cl.cam.ac.uk}
\authorinfo{Dominic Orchard}
           {Computer Laboratory, University of Cambridge, UK}
           {dominic.orchard\ensuremath{\mathord{@}}cl.cam.ac.uk}
\authorinfo{Alan Blackwell}
           {Computer Laboratory, University of Cambridge, UK}
           {alan.blackwell\ensuremath{\mathord{@}}cl.cam.ac.uk}

\title{Temporal semantics for a live coding language} 
% A programming model for temporal coordination (in music)}

@@

The \emph{Monad} instance for \emph{Temporal} is then as follows:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{Temporal}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{return}\;\Varid{a}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\mathbin{\char92 \char95 }\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{a},\Varid{vT})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{T}\;\Varid{p})\bind \Varid{q}{}\<[16]%
\>[16]{}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\Varid{nowT})\to \lambda \Varid{vT}\to {}\<[E]%
\\
\>[16]{}\hsindent{5}{}\<[21]%
\>[21]{}\mathbf{do}\;{}\<[25]%
\>[25]{}(\Varid{x},\Varid{vT'}){}\<[37]%
\>[37]{}\leftarrow \Varid{p}\;(\Varid{startT},\Varid{nowT})\;\Varid{vT}{}\<[E]%
\\
\>[25]{}\mathbf{let}\;(\Conid{T}\;\Varid{q'}){}\<[37]%
\>[37]{}\mathrel{=}\Varid{q}\;\Varid{x}{}\<[E]%
\\
\>[25]{}\Varid{thenT}{}\<[37]%
\>[37]{}\leftarrow \Varid{getCurrentTime}{}\<[E]%
\\
\>[25]{}\Varid{q'}\;(\Varid{startT},\Varid{thenT})\;\Varid{vT'}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
To ease understanding, we recall the types of \emph{return}
and \ensuremath{(\bind )} along with some intuition for their behaviour for
\emph{Temporal}:
%
\begin{itemize}
\item \ensuremath{\Varid{return}\mathbin{::}\Varid{a}\to \Conid{Temporal}\;\Varid{a}} lifts a pure value into a trivially
effectful computation by ignoring the time parameters and 
providing the usual pure state behaviour of returning the parameter state unchanged
(named \emph{vT} in this case).

@@

The following interpretation function $\interp{-}$ on \lang{} 
programs shows the mapping to the operations of the \emph{Temporal}
monad:
%%
\begin{align*}
\interp{\emph{statement}} & : \emph{Temporal} \, () \\
\interp{x = P; Q} & = \interp{P} \ensuremath{\bind (\lambda \Varid{x}\to } \interp{Q}) \\
\interp{P; Q} & = \interp{P} \ensuremath{\bind (\mathbin{\char92 \char95 }\to } \interp{Q}) \\
\interp{\sleep e} & = \emph{sleep} \, \interp{e}
\end{align*}
%%
Note that $\interp{-}$ is overloaded in the rule for \sleepOp{} for (pure) expressions. 
The concrete interpreation of other statements in the language, such as \playOp, is
elided here since it does not relate directly to the temporal semantics.

@@

\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{time}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
\\
\>[B]{}\Varid{time}{}\<[7]%
\>[7]{}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\Varid{nowT})\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{nowT},\Varid{vT})){}\<[E]%
\\
\>[B]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\Varid{start}\mathbin{::}\Conid{Temporal}\;\Conid{Time}{}\<[E]%
\\
\>[B]{}\Varid{start}\mathrel{=}\Conid{T}\;(\lambda (\Varid{startT},\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{startT},\Varid{vT})){}\<[E]%
\\
\>[B]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\Varid{getVirtualTime}\mathbin{::}\Conid{Temporal}\;\Conid{VTime}{}\<[E]%
\\
\>[B]{}\Varid{getVirtualTime}\mathrel{=}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to \Varid{return}\;(\Varid{vT},\Varid{vT})){}\<[E]%
\\
\>[B]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\Varid{setVirtualTime}\mathbin{::}\Conid{VTime}\to \Conid{Temporal}\;(){}\<[E]%
\\
\>[B]{}\Varid{setVirtualTime}\;\Varid{vT}\mathrel{=}\Conid{T}\;(\mathbin{\char92 \char95 }\to \mathbin{\char92 \char95 }\to \Varid{return}\;((),\Varid{vT})){}\<[E]%
\\
\>[B]{}\\[-1.5em]{}\<[E]%
\\
\>[B]{}\Varid{kernelSleep}\mathbin{::}\Conid{RealFrac}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{Temporal}\;(){}\<[E]%
\\
\>[B]{}\Varid{kernelSleep}\;\Varid{t}\mathrel{=}{}\<[18]%
\>[18]{}\Conid{T}\;(\lambda (\anonymous ,\anonymous )\to \lambda \Varid{vT}\to {}\<[E]%
\\
\>[18]{}\hsindent{6}{}\<[24]%
\>[24]{}\mathbf{do}\;{}\<[28]%
\>[28]{}\Varid{threadDelay}\;(\Varid{round}\;(\Varid{t}\mathbin{*}\mathrm{1000000})){}\<[E]%
\\
\>[28]{}\Varid{return}\;((),\Varid{vT})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Simple \emph{Temporal} computations, used  by the model}
\label{core-functions}
\end{figure}

@@

\subsection{Emitting overrun warnings}

We extend the \ensuremath{\Conid{Temporal}} monad with an additional parameter for the
$\epsilon$ time (maximum allowable overrun) and an output stream for
sending ``warnings'' when overruns occur.

Overrun warnings are either \emph{strong}, when the real time 
is more than $\epsilon$ ahead of virtual time, or \emph{weak} when the real
time is less than $\epsilon$ ahead of virtual time. That is:
%%
\begin{itemize}
\item{$\etime{P} > (\vtime{P} + \epsilon) \Rightarrow \interp{P} \leadsto$ \emph{strong warning} }
\item{$(\vtime{P} + \epsilon) \leq \etime{P} > \vtime{P} \Rightarrow \interp{P} \leadsto$ \emph{weak warning}}
\end{itemize}
%%

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Warning}\mathrel{=}\Conid{StrongOverrun}\;\Conid{VTime}\mid \Conid{WeakOverrun}\;\Conid{VTime}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{data}\;\Conid{TemporalE}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\Conid{VTime}\to \Conid{Temporal}\;(\Varid{a},[\mskip1.5mu \Conid{Warning}\mskip1.5mu])){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{Monad}\;\Conid{TemporalE}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{return}\;\Varid{a}\mathrel{=}\Conid{TE}\;(\mathbin{\char92 \char95 }\to \Varid{return}\;(\Varid{a},[\mskip1.5mu \mskip1.5mu])){}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}(\Conid{TE}\;\Varid{p})\bind \Varid{q}\mathrel{=}\Conid{TE}\;(\lambda \Varid{eps}\to \mathbf{do}\;(\Varid{a},\Varid{es})\leftarrow \Varid{p}\;\Varid{eps}{}\<[E]%
\\
\>[5]{}\hsindent{30}{}\<[35]%
\>[35]{}\mathbf{let}\;(\Conid{TE}\;\Varid{q'})\mathrel{=}\Varid{q}\;\Varid{a}{}\<[E]%
\\
\>[5]{}\hsindent{30}{}\<[35]%
\>[35]{}(\Varid{b},\Varid{es'})\leftarrow \Varid{q'}\;\Varid{eps}{}\<[E]%
\\
\>[5]{}\hsindent{30}{}\<[35]%
\>[35]{}\Varid{return}\;(\Varid{b},\Varid{es}\plus \Varid{es'})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
